An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
Work out the encoding

@@

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid \Fudge{MONOID} and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

@@

\contents{\item Monoid
\item Alfabet
\item ord oever alfabet
\item konkatenering
}

\subsection{Finite-state Automata and their Products}
We define a non-deterministic automaton~$\Automaton$ with alphabet~$\Alphabet$
as $\Automaton = \AutomatonTuple$ where $\Transitions = \States \times \Alphabet
\times \States$, $\States$ is its states, $\InitialState$ is w.l.o.g. assumed to
be the single initial state, and $\AcceptingStates$ is its set of accepting
states.  We describe a transition $\Transition$ from state $\State$ to state
$\State'$ with label $\Label \in \Alphabet$ as $\Transition =
\FromLabelTo{\State}{\Label}{\State'}$. \Fudge{When one or more of the states and
labels of a transition are uninteresting we will omit them}.

@@

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l(w)$ to talk about an individual letter $l$ in a word $w$. For
example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have
already seen this in Section~\ref{sec:introduction:parikh}.

Parikh's theorem states that any context-free language has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \CountOf{a} =
\CountOf{b} = \CountOf{c} \land \CountOf{a} \geq 0$).

\contents{\item this means we can represent this as Presburger
\item only interested in regular languages}

@@

To generalise the logic of the Parikh map, we introduce a morphism $\Map:\: \Sigma^*
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$, where $\Concat$ is
  the string concatenation operation
\end{itemize}

Many other generalisations of the Parikh map have been studied. Prominent
examples include generalising the Parikh map to segments of a fixed length
\cite{KARHUMAKI1980155} and the more general Parikh matrix, which gives more
information about a word than the standard Parikh image \cite{parikh-matrix}.
Another notable generalisation is the p-vector, introduced
in~\cite{infinite-words}, which denotes the position of each letter in the word
rather than the number of their occurrences and allows for generalisations into
infinite alphabets. All of these in some sense extend the Parikh map. By
contrast, the main utility of the formulation introduced here is that it allows
us to \emph{remove} something, thereby potentialy obtaining an easier problem.

@@

Since Parikh images are semilinear, they can be fully expressed by Presburger
arithmetic. This means that any Parikh image can be written as a set of linear
equation. Following~\cite{generate-parikh-image}, we define the Parikh Image of
a regular language recognised by a DFA $\Automaton = \AutomatonTuple$ as:

\contents{\item replace implies with singlearrow
\item splitta sista implikationerna i tvaa!
\item intuition for what the variables mean}

\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \implies \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} \Filter(\Transition) - \sum_{\Transition = \FromLabelTo{}{}{\State}} \Filter(\Transition)
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \implies \StateVar_{\State} > 0
} \\
& \AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \implies
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates \land \right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

where all variables $\TransitionVar_i, \StateVar_i, \FinalStateVar_i$ are
existentially quantified and the free variables $\LetterVar_\Letter$ make up the
actual image. Performing standard quantifier elimination on $\TransitionVar_i,
\StateVar_i, \FinalStateVar_i$ with $\LetterVar_\Letter$ would produce the
Parikh image in its Presburger form. Note that compared to the method described
in~\cite{generate-parikh-image}, ours is backwards. Intuitively the approaches
are equivalent since the Parikh image, by definition, disregards order. The
approach can be described as either following paths through the automaton
forwards from an (in our case, \emph{the}) initial state, or conversely as in
the case here, choose an accepting state and trace a path backwards to the
initial state.

In this paper we refer to this model as the baseline approach, though we also
apply some optimisations that allow us to detect unsatisfiability early,
described in Section~\ref{sec:implementing-baseline}. Notably, the calculus
introduced in this paper can at its core be described as lazily enforcing the
connectedness constraint of this method (the final two clauses) while also
interleaving the (similarly lazy) computation of products of automata.

@@

Note that one of the $\lor$-clauses can be immediately removed since it
contains the contradiction $\StateVar_{1} = \StateVar_{1} + 1$, arising from
cycles. These clauses encode the distance in a spanning tree from an accepting
state, and the unsatisfiability is by design, essentially ensuring that two
states in a cycle cannot vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter a, i.e. we solve
for $\LetterVar_a > 0$. This implies that $\StateVar_{1} > 0 \land \StateVar_{3}
> 0$ (essentially, that the state our transition of interest started in is
reachable), in turn implying that $\FinalStateVar_3 = 1$. Plugging in these
facts, we get:

\begin{equation}
  \begin{aligned}
  &\LetterVar_{a} = 2 \land \LetterVar_{b} = 0 \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}}\\
  & \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula-calc}
  \end{equation}

>>>>

Many other generalisations of the Parikh than the ones we use here have been
studied. Prominent examples include generalising the Parikh map to segments of a
fixed length \cite{KARHUMAKI1980155} and the more general Parikh matrix, which
gives more information about a word than the standard Parikh image
\cite{parikh-matrix}. Another notable generalisation is the p-vector, introduced
in~\cite{infinite-words}, which denotes the position of each letter in the word
rather than the number of their occurrences and allows for generalisations into
infinite alphabets. All of these in some sense extend the Parikh map. By
contrast, the main utility of the formulation introduced here is that it allows
us to \emph{remove} something, thereby potentialy obtaining an easier problem.

@@

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

@@

\subsection{Monoids}

We write commutative monoids as $\Monoid =
\left(X;+_{\Monoid};0_{\Monoid}\right)$, where we mean $X$ to be the underlying
set of elements, $+_{\Monoid}$ to be the (commutative) monoid map, with
$0_{\Monoid}$ as its neutral element.

\subsection{Languages, Finite-state Automata and their Products}

We define an alphabet as a set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.

A non-deterministic automaton~$\Automaton$ with alphabet~$\Alphabet$ is
$\AutomatonTuple$ where $\Transitions = \States \times \Alphabet \times
\States$, $\States$ is its states, $\InitialState$ is w.l.o.g. assumed to be the
single initial state, and $\AcceptingStates$ is its set of accepting states.  We
write a transition $\Transition = \Tuple{\State, \Label, \State'} \in
\Transitions$ as $\Transition = \FromLabelTo{\State}{\Label}{\State'}$.
Similarly, we use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to refer to the
set of transitions coming into $\State$, whenever the automaton is clear from
the context.

@@

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l(w)$ to talk about an individual letter $l$ in a word $w$. For
example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have
already seen this in \cref{sec:introduction:parikh}.

Parikh's theorem states that any context-free language has a letter-equivalent
regular language (c.f.~\cite{construction} for a construction of such automata
from context-free grammars and~\cite{bounds} for bounds on its size). However,
there are languages that are not context-free that also have semilinear images
under~$\ParikhMap$ (e.g. $\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) =
\ParikhMap((abc)^*) = \CountOf{a} = \CountOf{b} = \CountOf{c} \land \CountOf{a}
\geq 0$). This means they can be represented as a quantifier-free Presburger
formula.

Note that while Parikh's theorem applies to context-free languages, in this
paper we are only interested in regular languages.

@@

To generalise the logic of the Parikh map, we introduce a morphism $\Map:\: \Sigma^*
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$
\end{itemize}

@@

Since Parikh images are semilinear, any Parikh image can be written as a set of linear
equations. The following construction, here adapted to work on an NFA $\Automaton = \AutomatonTuple$, was presented in\cite{generate-parikh-image}:
\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \longrightarrow  \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition = \FromLabelTo{\State}{}{}} \Filter(\Transition) - \sum_{\Transition = \FromLabelTo{}{}{\State}} \Filter(\Transition)
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{\State'}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \longrightarrow \StateVar_{\State} > 0
} \\
& \AndComp{\State \in F}{
  \StateVar_{\State} > 0 \longrightarrow \FinalStateVar_{\State} = 1
} \\
&\AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \longrightarrow
  \left(\FinalStateVar_{\State} = 1 \text{ if } \State \in \AcceptingStates \land \right) \OrComp{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

All variables $\TransitionVar_i, \StateVar_i, \FinalStateVar_i$ are
existentially quantified and the free variables $\LetterVar_\Letter$ make up the
image. $\StateVar_\State$ represents the distance of state $\State$ from
$\InitialState$ in a spanning tree, $\TransitionVar_\Transition$ how many times
a transition $\Transition$ is used, and $\FinalStateVar_\State$ whether a given
accepting state $\State \in \AcceptingStates$ is the actually used final state.

In this paper we refer to this model as the baseline approach, though we also
apply optimisations as described in \cref{sec:implementing-baseline}. The
calculus introduced in this paper, by contrast, lazily enforces the
connectedness constraint of this encoding (the final three clauses) while also
interleaving the computation of products of automata and propagating information
between the steps to reduce the amount of work that needs to be done.

@@

Note that one of the $\lor$-clauses can be immediately removed since it contains
the contradiction $\StateVar_{1} = \StateVar_{1} + 1$, arising from cycles.The
unsatisfiability is by design, and ensures that two states in a cycle cannot
vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter a, i.e. we solve
for $\LetterVar_a > 0$. This implies that $\StateVar_{1} > 0 \land \StateVar_{3}
> 0$ (essentially, that the state our transition of interest started in is
reachable), in turn implying that $\FinalStateVar_3 = 1$. Plugging in these
facts, we get:
\begin{equation}
  \begin{aligned}
  &\LetterVar_{a} = 2 \land \LetterVar_{b} = 0 \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}}\\
  & \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula-calc}
  \end{equation}
